---
title: "Single-Cell RNA-seq Analysis of Lead-Exposed Human Neural Progenitor Cells"
subtitle: "CAJAL Advanced Neuroscience Course — Bordeaux"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)
```

------------------------------------------------------------------------

# Before You Begin: Required Files

All of the following files must be present in your working directory
before running this worksheet. If any are missing, ask your instructor.

| File                             | Description                                                  |
|--------------------------|----------------------------------------------|
| `downsampled_Pb40.qs`            | Downsampled Seurat object (\~100k cells, pre-QC)             |
| `cell_metadata.csv`              | Barcode-to-donor demultiplexing output from Dropulation      |
| `ensembl_to_symbol.tsv`          | Gencode V41 Ensembl ID to gene symbol and chromosome mapping |
| `cell_cycle_markers_ensembl.csv` | S and G2M phase gene sets in Ensembl ID format               |
| `NPCvsOther_genelist.csv`        | Marker genes distinguishing NPCs from other cell types       |
| `NPCmarkers_shortgenelist.csv`   | Curated shortlist of canonical NPC marker genes              |

All intermediate objects generated during the analysis will be saved to
`Analysis_Outputs/0_seurat-objects/` in your working directory.

------------------------------------------------------------------------

# Section 1: Biological and Experimental Context

## 1.1 The SNaP40 Pb Village Dataset

The data you will analyze comes from an experiment designed to
understand how **environmental lead (Pb) exposure affects human neural
progenitor cell (NPC) transcription** across a genetically diverse
population.

NPCs are the proliferating progenitors that give rise to neurons and
glia during brain development. They are particularly relevant for
studying neurotoxicant exposure because the developing brain is far more
sensitive to environmental insults than the adult brain — and because
NPC dysfunction during development has lasting consequences for neuronal
circuit formation and cognitive function.

The cells in this experiment were derived from **induced pluripotent
stem cells (iPSCs)** reprogrammed from 40 human donors with diverse
genetic backgrounds. These iPSCs were differentiated into NPCs and then
treated with lead acetate at three concentrations:

-   **Control** — no lead exposure
-   **Low** — 3 µM lead acetate
-   **High** — 10 µM lead acetate

This experimental design is called a **"village" experiment**: rather
than differentiating and treating each donor's cells independently
(which would be prohibitively expensive and introduce batch effects
between donors), all 40 donors' cells are pooled together into a single
culture dish per condition and treated simultaneously. This is possible
because each donor's cells carry a unique genotype, allowing them to be
computationally identified and separated after sequencing using a
process called **genotype-based demultiplexing**.

## 1.2 Why 40 Donors, and Why It Matters for Analysis

By profiling 40 donors simultaneously, the experiment has enough
statistical power to ask: *what is the shared transcriptional response
to lead across diverse human genetic backgrounds?*

This design has a direct and important consequence for how we analyze
the data. Each donor contributes thousands of cells to the dataset.
Those cells are not independent observations — they share a genome, an
epigenome, and a differentiation history. If we treat every cell as
independent when testing for differential expression, we will massively
overstate our statistical confidence. The correct approach is to treat
**donor as the unit of replication**, not individual cells. We will
return to this point in detail when we reach the differential expression
section.

## 1.3 What This Worksheet Covers

This worksheet walks you through the complete analysis pipeline:

1.  Loading the data and attaching donor metadata
2.  Quality control — identifying and removing low-quality cells
3.  Normalization, dimensionality reduction, and clustering
4.  Assessing potential confounders (batch effects, cell cycle)
5.  Cell type annotation using marker genes
6.  Pseudobulk differential expression with `dreamlet`
7.  Visualization of DEG results
8.  Gene set enrichment analysis with `fgsea`

------------------------------------------------------------------------

# Section 2: Setup and Data Loading

## 2.1 Libraries

We load all required libraries upfront. If any package is not installed,
use `install.packages()` for CRAN packages or `BiocManager::install()`
for Bioconductor packages.

```{r load-libraries, message=FALSE, warning=FALSE}
# ── Project file path management ──────────────────────────────────────────────
library(here)           # here() anchors paths to the R project root
here::i_am("scrnaseq_analysis.Rproj")

# ── Data wrangling ─────────────────────────────────────────────────────────────
library(tidyverse)      # dplyr, ggplot2, readr, and friends
library(data.table)     # fread() for fast flat file reading
library(skimr)          # skim() for quick data summaries

# ── Single-cell analysis ───────────────────────────────────────────────────────
library(Seurat)         # Core scRNA-seq analysis toolkit (v5)
library(Matrix)         # Sparse matrix support (~90% of scRNA-seq entries are 0)
library(qs)             # qsave()/qread() — faster than saveRDS()/readRDS()
library(scCustomize)    # QC plotting helpers and scRNA-seq utilities
library(scater)         # Additional QC and single-cell utilities

# ── Visualization ──────────────────────────────────────────────────────────────
library(patchwork)      # Combining multiple ggplot objects
library(ggrepel)        # Non-overlapping text labels on plots
library(viridis)        # Colorblind-friendly color scales
```

```{r setup-directories}
# Create a directory to store all intermediate Seurat objects.
# All saved files will go here throughout the worksheet.
output_dir <- here("Analysis_Outputs", "0_seurat-objects")

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  message("Created output directory: ", output_dir)
} else {
  message("Output directory already exists: ", output_dir)
}
```

## 2.2 Loading the Seurat Object

The raw data for this experiment was aligned and quantified using
STARsolo, which outputs per-cell count matrices with one row per gene
and one column per cell barcode. These matrices were merged across the
six experimental runs (two per condition) into a single Seurat object,
and the full dataset of \~400,000 cells was downsampled to \~100,000
cells for this course.

A **Seurat object** is an S4 R object that stores count matrices,
metadata, dimensionality reductions, and cluster assignments in a single
place. When we first load it, it contains only the raw count matrix and
three default metadata columns: `orig.ident` (which sample the cell came
from), `nCount_RNA` (total UMIs detected), and `nFeature_RNA` (number of
unique genes detected). Donor identity, treatment condition, and other
biological metadata have not yet been attached.

```{r load-seurat-object}
# qread() is the companion to qsave() — both come from the qs package.
# It is substantially faster than readRDS() for large R objects.
gex <- qread(here("downsampled_Pb40.qs"))

# Check the dimensions: rows = genes, columns = cells
dim(gex)

# Inspect the metadata — at this stage only three columns exist
head(gex@meta.data)

# How many cells came from each experimental run?
table(gex$orig.ident)
```

## 2.3 Adding Donor-Level Metadata

Because this is a village experiment, cells from all 40 donors are
pooled together in each condition. After sequencing, each cell barcode
is assigned to its donor of origin using **Dropulation**, a
genotype-based demultiplexer. Dropulation compares the SNPs present in
each cell's reads to a reference panel of known donor genotypes and
assigns each cell barcode a probability-based donor identity.

The output of this demultiplexing step — one row per cell barcode, with
donor identity, treatment condition, sex, and other covariates — has
been prepared for you in `cell_metadata.csv`. We attach it to the Seurat
object using `AddMetaData()`, which joins the table by matching the
`cellid` column to cell barcodes in `gex`.

```{r add-donor-metadata}
# Load the demultiplexing output and set cell barcodes as rownames
# so AddMetaData() can match them to the Seurat object
cell_metadata <- read_csv(here("cell_metadata.csv")) %>%
  column_to_rownames("cellid")

# Attach to the Seurat object
gex <- AddMetaData(gex, cell_metadata)

# Confirm the join worked — you should now see donor, treatment, sex,
# origin, donorlabel, and sample_index alongside the original columns
colnames(gex@meta.data)
```

```{r inspect-metadata}
# How many cells were successfully assigned to a donor?
# Cells with NA could not be confidently assigned and will be removed later.
table(is.na(gex$donor))

# Cells per treatment condition — should be roughly balanced
table(gex$treatment)

# Cells per donor — notice this is intentionally uneven. Genotype-based
# demultiplexing assigns cells probabilistically, and some donors will be
# better represented than others. This imbalance is real and will matter
# when we aggregate to pseudobulk later: donors with very few cells in a
# given cell type may be dropped from that cell type's model.
table(gex$donor)

# Save the annotated object
qsave(gex, file.path(output_dir, "00-gex_with_metadata.qs"))
```

------------------------------------------------------------------------

# Section 3: Quality Control

## 3.1 Why QC Matters

In scRNA-seq, each oligo tagged bead ideally captures the DNA from a
single cell. In practice, a meaningful fraction of captured beads
contain something other than a healthy intact cell:

-   **Empty beads**— the bead captured ambient RNA but no cell. These
    tend to have very low UMI counts and few detected genes.
-   **Damaged or dying cells** — the cell membrane has ruptured before
    capture, causing cytoplasmic mRNA to leak out. What remains is
    enriched for mitochondrial transcripts (which are protected inside
    mitochondria), so these barcodes have a disproportionately high
    fraction of mitochondrial reads.
-   **Doublets** — two cells were captured in the same well. These tend
    to have unusually high UMI counts and gene counts relative to the
    rest of the dataset.

We use four per-cell metrics to identify and remove these low-quality
barcodes:

-   **`nCount_RNA`** — total UMIs detected. Very low = empty bead; very
    high relative to the distribution = potential doublet.
-   **`nFeature_RNA`** — number of unique genes detected. Low values
    indicate low-complexity barcodes.
-   **`percent_mito`** — fraction of UMIs mapping to mitochondrial
    genes. High values indicate damaged or lysed cells.
-   **`log10GenesPerUMI`** — transcriptional complexity score
    (log10(nFeature) / log10(nCount)). Damaged cells and doublets tend
    to have lower complexity than healthy single cells.

An important principle: these metrics should be assessed **jointly**,
not independently. A cell with moderately elevated mitochondrial
fraction but high UMI count is less concerning than one that fails all
four metrics simultaneously.

## 3.2 Computing QC Metrics

Our genes are stored as **Ensembl IDs** (e.g., `ENSG00000141510`) rather
than gene symbols (e.g., `MT-CO1`). This means we cannot use simple
pattern matching to identify mitochondrial genes — we need to use a
genome annotation file to look up which Ensembl IDs correspond to genes
encoded on the mitochondrial chromosome.

```{r compute-qc-metrics}
# Load the Gencode V41 annotation file.
# This maps Ensembl gene IDs to gene symbols, chromosome (seqnames),
# gene type, and other annotation fields.
ensg_to_symbol <- read_tsv(here("ensembl_to_symbol.tsv"))

# Ensembl IDs in annotation files often carry version numbers
# (e.g., ENSG00000141510.18). Strip the version suffix so they match
# the IDs stored in our Seurat object.
ensg_to_symbol <- ensg_to_symbol %>%
  mutate(gene_id = gsub("\\..*", "", gene_id))

# Mitochondrial genes are encoded on chrM. Pull their Ensembl IDs.
genes_MT <- ensg_to_symbol %>%
  filter(seqnames == "chrM") %>%
  pull(gene_id) %>%
  unique()

message(sprintf("%d mitochondrial genes identified", length(genes_MT)))

# scCustomize provides a curated list of human ribosomal gene Ensembl IDs
# (ensembl_ribo_id$Homo_sapiens_ribo_ensembl). Cross-reference against our
# annotation to get only those present in this dataset.
genes_ribo <- ensg_to_symbol %>%
  filter(gene_id %in% ensembl_ribo_id$Homo_sapiens_ribo_ensembl) %>%
  pull(gene_id) %>%
  unique()

message(sprintf("%d ribosomal genes identified", length(genes_ribo)))

# Add_Mito_Ribo() calculates percent_mito and percent_ribo for each cell
# and appends them to gex@meta.data. We pass our Ensembl-based lists
# explicitly — the default symbol-based detection would fail here.
gex <- Add_Mito_Ribo(
  gex,
  species       = "Human",
  mito_features = genes_MT,
  ribo_features = genes_ribo
)

# Add_Cell_Complexity() computes log10(nFeature_RNA) / log10(nCount_RNA)
# and stores it as log10GenesPerUMI
gex <- Add_Cell_Complexity(gex)

# Confirm the new columns are present
colnames(gex@meta.data)
```

```{r skim-qc-metrics}
# A quick numerical summary of all QC metrics across all cells.
# Pay attention to the distributions of percent_mito and log10GenesPerUMI —
# these will guide threshold selection below.
skim(gex@meta.data %>% select(nCount_RNA, nFeature_RNA,
                               percent_mito, percent_ribo,
                               log10GenesPerUMI))
```

## 3.3 Visualizing QC Distributions

Before applying any filter, look at the data. Thresholds should be
chosen based on the distributions you actually observe — not copied from
a paper using a different cell type, protocol, or sequencing depth.

```{r qc-summary-table}
# Median QC statistics per sample. Record these — they belong in your
# methods section when you write this up.
(prefilter_stats <- Median_Stats(seurat_object = gex, group_by_var = "orig.ident"))
```

```{r set-thresholds}
# Define your QC thresholds here. These variables are used in all
# subsequent plots and in the filtering step. Adjust and re-run the
# visualization chunks to see how different values would change your
# retained cell count before committing.
low_umi        <- 1000
low_feature    <- 500
high_mito      <- 25
low_complexity <- 0.85
```

```{r plot-qc-violins, fig.width=12, fig.height=8}
# Violin plots per sample. The dashed reference lines show where your
# current thresholds sit — they do NOT filter cells at this step.
# Look for consistency across samples: if one sample has a dramatically
# different distribution, investigate before applying a global threshold.
p1 <- QC_Plots_UMIs(gex,       low_cutoff  = low_umi,        pt.size = 0, plot_boxplot = TRUE)
p2 <- QC_Plots_Genes(gex,      low_cutoff  = low_feature,    pt.size = 0, plot_boxplot = TRUE)
p3 <- QC_Plots_Mito(gex,       high_cutoff = high_mito,      pt.size = 0, plot_boxplot = TRUE)
p4 <- QC_Plots_Complexity(gex, low_cutoff  = low_complexity, pt.size = 0, plot_boxplot = TRUE)

wrap_plots(p1, p2, p3, p4, ncol = 2)

ggsave(here("Analysis_Outputs", "3-2_qc_violins_prefilter.png"),
       width = 12, height = 8, dpi = 300)

```

```{r plot-qc-scatter, fig.width=12, fig.height=10}
# Scatter plots show how metrics relate to each other across individual cells.
# A cell that fails multiple metrics simultaneously is a much more confident
# removal candidate than one that barely crosses a single threshold.
meta_data <- Fetch_Meta(gex)

# Genes vs UMIs, colored by % mito.
# Healthy cells follow a tight diagonal — more sequencing depth (UMIs)
# yields more detected genes. Cells sitting below the diagonal have fewer
# unique genes than expected for their depth. Notice that the high-mito
# cells (warm colors) tend to cluster in the lower-left: low UMIs, few
# genes, high mito is the classic signature of a lysed cell.
# Also notice the geometric "shelf" of points with moderate-to-high UMIs
# but elevated mito — these are likely doublets, where one healthy cell
# and one damaged cell were co-captured in the same bead

p1 <- ggplot(meta_data, aes(x = nCount_RNA, y = nFeature_RNA, color = percent_mito)) +
  geom_point(size = 0.3, alpha = 0.4) +
  scale_color_viridis_c(option = "plasma") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = low_umi,     linetype = "dashed", color = "red") +
  geom_hline(yintercept = low_feature, linetype = "dashed", color = "red") +
  labs(title = "Genes vs UMIs (colored by % mito)",
       x = "log10(UMIs per cell)", y = "log10(Genes per cell)",
       color = "% Mito") +
  theme_classic()

# UMIs vs % mito.
# High mito + low UMIs is the clearest indicator of a damaged cell.
# The bimodal structure you see in the density plot below is visible here
# too: a dense healthy population hugging the x-axis, and a high-mito
# cloud spanning a wide range of UMI counts.
p2 <- ggplot(meta_data, aes(x = nCount_RNA, y = percent_mito, color = nFeature_RNA)) +
  geom_point(size = 0.3, alpha = 0.4) +
  scale_color_viridis_c(option = "plasma") +
  scale_x_log10() +
  geom_hline(yintercept = high_mito, linetype = "dashed", color = "red") +
  labs(title = "UMIs vs % Mitochondrial Expression",
       x = "log10(UMIs per cell)", y = "% Mitochondrial",
       color = "Genes") +
  theme_classic()

# Complexity vs % mito.
# The upper-left quadrant — low complexity and high mito — contains the
# most confidently low-quality cells. Cells in the lower-right are your
# healthy population.
p3 <- ggplot(meta_data, aes(x = log10GenesPerUMI, y = percent_mito,
                              color = orig.ident)) +
  geom_point(size = 0.3, alpha = 0.4) +
  geom_vline(xintercept = low_complexity, linetype = "dashed", color = "red") +
  geom_hline(yintercept = high_mito,      linetype = "dashed", color = "red") +
  labs(title = "Complexity vs % Mitochondrial Expression",
       x = "log10GenesPerUMI", y = "% Mitochondrial",
       color = "Sample") +
  theme_classic()

# Density of % mito per sample.
# This is where the bimodal distribution is most apparent. The sharp peak
# around 5-10% represents intact healthy cells. The broad shoulder toward
# 75% reflects damaged cells and doublets. A well-chosen mito threshold
# should sit in the valley between these two modes — which is what our
# 25% cutoff does. All six samples overlap closely, which tells us the
# library quality is consistent across conditions.
p4 <- ggplot(meta_data, aes(x = percent_mito, fill = orig.ident)) +
  geom_density(alpha = 0.3) +
  geom_vline(xintercept = high_mito, linetype = "dashed", color = "red") +
  labs(title = "% Mitochondrial Expression per Sample",
       x = "% Mitochondrial", y = "Density") +
  theme_classic()

wrap_plots(p1, p2, p3, p4, ncol = 2)
ggsave(here("Analysis_Outputs", "3-2_qc_scatter_prefilter.png"),
       width = 12, height = 10, dpi = 300)
```

## 3.4 Filtering

We apply two sequential filters. First, we remove cells with no donor
assignment — these cannot contribute to our donor-level differential
expression model. About 19% of barcodes fall into this category, which
is normal for genotype-based demultiplexing and reflects barcodes that
could not be confidently assigned to a single donor.

Second, we apply the QC metric thresholds. A cell must pass **all four**
criteria simultaneously to be retained.

Before committing, use the exploration block below to see exactly how
many cells each threshold combination would retain.

```{r explore-thresholds}
# Test your thresholds without modifying gex.
# Change low_umi, low_feature, high_mito, or low_complexity in the
# set-thresholds chunk above and re-run this block to see the effect.
test_subset <- subset(
  x = gex[, !is.na(gex$donor)],
  subset = (nFeature_RNA     >= low_feature)    &
           (nCount_RNA       >= low_umi)         &
           (log10GenesPerUMI >= low_complexity)  &
           (percent_mito     <  high_mito)
)

message(sprintf(
  "Thresholds: UMIs >= %d | genes >= %d | complexity >= %.2f | mito < %d%%",
  low_umi, low_feature, low_complexity, high_mito
))
message(sprintf(
  "Cells retained: %d of %d donor-assigned cells (%.1f%%)",
  ncol(test_subset),
  sum(!is.na(gex$donor)),
  100 * ncol(test_subset) / sum(!is.na(gex$donor))
))
```

```{r apply-qc-filters}
# Step 1: Remove unassigned cells
message(sprintf("Total cells before any filtering: %d", ncol(gex)))

gex <- gex[, !is.na(gex$donor)]

message(sprintf("After removing NA-donor cells: %d", ncol(gex)))

# Step 2: Apply QC thresholds using the variables defined above
gex_subset <- subset(
  x = gex,
  subset = (nFeature_RNA     >= low_feature)    &
           (nCount_RNA       >= low_umi)         &
           (log10GenesPerUMI >= low_complexity)  &
           (percent_mito     <  high_mito)
)

message(sprintf("After QC filtering: %d cells retained", ncol(gex_subset)))
message(sprintf(
  "Removed by QC: %d cells (%.1f%%)",
  ncol(gex) - ncol(gex_subset),
  100 * (1 - ncol(gex_subset) / ncol(gex))
))
```

```{r plot-postfilter-qc, fig.width=12, fig.height=8}
# Compare to the pre-filter violins above. The low-quality tails should
# be substantially reduced — but the main body of the distribution should
# look similar, confirming we haven't over-filtered.
p1 <- QC_Plots_UMIs(gex_subset,       low_cutoff  = low_umi,        pt.size = 0, plot_boxplot = TRUE)
p2 <- QC_Plots_Genes(gex_subset,      low_cutoff  = low_feature,    pt.size = 0, plot_boxplot = TRUE)
p3 <- QC_Plots_Mito(gex_subset,       high_cutoff = high_mito,      pt.size = 0, plot_boxplot = TRUE)
p4 <- QC_Plots_Complexity(gex_subset, low_cutoff  = low_complexity, pt.size = 0, plot_boxplot = TRUE)

wrap_plots(p1, p2, p3, p4, ncol = 2)
ggsave(here("Analysis_Outputs", "3-4_qc_violins_postfilter.png"),
       width = 12, height = 8, dpi = 300)

```

```{r plot-cells-per-sample, fig.width=7, fig.height=4}
# Cells remaining per sample after filtering.
# A large discrepancy between samples would warrant investigation —
# it could mean one library had systematically worse quality.
Fetch_Meta(gex_subset) %>%
  ggplot(aes(x = orig.ident, fill = orig.ident)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5, size = 3.5) +
  theme_classic() +
  theme(axis.text.x    = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none") +
  labs(title = "Cells per Sample after QC Filtering",
       x = NULL, y = "Number of Cells")

ggsave(here("Analysis_Outputs", "3-4_cells_per_sample_postfilter.png"),
       width = 7, height = 4, dpi = 300)

# Save the filtered object
qsave(gex_subset, file.path(output_dir, "01-gex_QC_filtered.qs"))
```

------------------------------------------------------------------------

# Section 4: Normalization, Dimensionality Reduction, and Clustering

## 4.1 Normalization and Scaling

Raw UMI counts are not directly comparable across cells because cells
differ in how many total transcripts were captured — partly due to real
differences in cell size and transcriptional activity, and partly due to
technical variation in sequencing depth. A gene with 100 counts in a
cell that has 5,000 total UMIs represents a very different expression
level than 100 counts in a cell with 50,000 total UMIs.

We apply three sequential transformations to make cells comparable:

**Normalization** scales each cell's counts to a common total of 10,000
UMIs, then log-transforms (`log1p`). This puts all cells on a comparable
scale and compresses the dynamic range of highly expressed genes.

**Variable feature selection** identifies the 2,000 genes with the
highest cell-to-cell variation. Running PCA on all \~20,000 genes would
add noise from housekeeping genes expressed at similar levels in every
cell — restricting to variable features focuses the dimensionality
reduction on genes that actually differentiate cell states.

**Scaling** centers each gene to zero mean and unit variance across
cells. This ensures that genes with high absolute expression do not
dominate the PCA simply because of their scale.

```{r normalize-scale-pca}
# Free memory from the pre-filter object before the compute-heavy steps
rm(gex)
gc()

# Log-normalize to 10,000 UMIs per cell
gex_subset <- NormalizeData(gex_subset, normalization.method = "LogNormalize",
                             scale.factor = 10000)

# Identify the 2,000 most variable genes
gex_subset <- FindVariableFeatures(gex_subset, nfeatures = 2000)

# Scale — by default scales all genes, not just variable ones, which is
# important for visualization later
gex_subset <- ScaleData(gex_subset)

pca_npcs = 30

# Run PCA on the variable features
gex_subset <- RunPCA(gex_subset, npcs = 30)

# The ElbowPlot shows variance explained per PC. You are looking for the
# "elbow" — the point at which additional PCs explain diminishing variance.
# The PCs up to and including the elbow will be used for clustering and UMAP.
ElbowPlot(gex_subset, ndims = 30)

ggsave(here("Analysis_Outputs", "4-1_pca_elbowplot.png"),
       width = 6, height = 4, dpi = 300)

```

```{r pca-feature-loadings, fig.width=10, fig.height=6}
# Inspect what drives the top PCs. VizDimLoadings shows the genes with the
# highest positive and negative loadings for each PC, which gives you a
# sense of what biology each component is capturing.

rownames(gex_subset@reductions$pca@feature.loadings) <- 
  ensg_to_symbol$gene_name[match(
    rownames(gex_subset@reductions$pca@feature.loadings),
    ensg_to_symbol$gene_id
  )]

# Any IDs that didn't match will become NA — replace those with original ID
na_idx <- is.na(rownames(gex_subset@reductions$pca@feature.loadings))
rownames(gex_subset@reductions$pca@feature.loadings)[na_idx] <- 
  rownames(gex_subset@reductions$pca@feature.loadings.projected)[na_idx]

VizDimLoadings(gex_subset, dims = 1:4, reduction = "pca", ncol = 4)
ggsave(here("Analysis_Outputs", "4-1_pca_feature_loadings.png"),
       width = 10, height = 6, dpi = 300)
```

## 4.2 Cell Cycle Scoring

NPC cultures are proliferating, which means a significant fraction of
cells will be actively cycling at any given time. Actively cycling cells
can cluster by cell cycle phase rather than by cell identity, which
would obscure the biology we care about. We score every cell for
expression of canonical S-phase and G2/M-phase gene sets to quantify
this effect.

The scores are added as covariates here and will become relevant when we
decide whether to regress out the cell cycle signal in the next section.

```{r cell-cycle-scoring}
# Load the curated cell cycle gene sets (Ensembl IDs)
cell_cycle_markers <- read_csv(here("cell_cycle_markers_ensembl.csv"))

s.genes   <- cell_cycle_markers %>% filter(phase == "S")    %>% pull(geneID)
g2m.genes <- cell_cycle_markers %>% filter(phase == "G2/M") %>% pull(geneID)

# CellCycleScoring() assigns each cell an S.Score, G2M.Score, and a
# predicted Phase (G1, S, or G2M) based on expression of the gene sets.
DefaultAssay(gex_subset) <- "RNA"
gex_subset <- CellCycleScoring(gex_subset,
                                s.features   = s.genes,
                                g2m.features = g2m.genes,
                                set.ident    = TRUE)

# What fraction of cells are in each phase?
table(gex_subset$Phase)
```

## 4.3 Clustering and UMAP

With PCA complete, we build a nearest-neighbor graph in PC space and
apply the Louvain algorithm to identify clusters of transcriptionally
similar cells. We then use UMAP to project the high-dimensional PC
embedding into two dimensions for visualization.

Two parameters govern clustering:

-   **`dims`** — how many PCs to use, chosen from the elbow plot above.
    We use 7 here based on the original analysis.
-   **`resolution`** — controls cluster granularity. Higher values
    produce more, smaller clusters. Start low (0.1–0.3) and increase if
    you find clusters that seem to contain distinct biology.

```{r cluster-and-umap}
npcs <- 26  # Chosen from the elbow plot — adjust if your elbow looks different

gex_subset <- gex_subset %>%
  FindNeighbors(dims = 1:npcs) %>%
  FindClusters(resolution = 0.2) %>%
  RunUMAP(dims = 1:npcs, return.model = TRUE)
  # return.model = TRUE preserves the UMAP model for projecting new data later

# How many cells are in each cluster?
table(gex_subset$seurat_clusters)
```

```{r plot-initial-umap, fig.width=12, fig.height=5}
# UMAP colored by cluster identity
p1 <- DimPlot(gex_subset, group.by = "seurat_clusters", label = TRUE, label.size = 4,
              reduction = "umap") + NoLegend() +
  labs(title = "Clusters")

# UMAP colored by cell cycle phase — generated immediately so you can
# assess whether phase structure aligns with cluster structure.
# If cells are clustering by phase rather than cell type, regression
# may be warranted (Section 5)
p2 <- DimPlot(gex_subset, group.by = "Phase",
              reduction = "umap",
              shuffle = TRUE, seed = 42) +
  labs(title = "Phase")

p1 + p2

ggsave(here("Analysis_Outputs", "4-3_initial_umap.png"),
       width = 12, height = 5, dpi = 300)
```

```{r plot-qc-on-umap, fig.width=12, fig.height=5}
# Overlay QC metrics on the UMAP. If filtering worked well, you should
# not see strong gradients in mito% or nCount across the embedding.
# Residual quality structure (e.g., a cluster with high mito) would
# suggest those cells survived filtering but are still suspect.
FeaturePlot(gex_subset,
            features  = c("nCount_RNA", "percent_mito"),
            reduction = "umap",
            order     = TRUE,
            ncol      = 2)

ggsave(here("Analysis_Outputs", "4-3_qc_umap.png"),
       width = 12, height = 5, dpi = 300)
```

```{r save-clustered-object}
qsave(gex_subset, file.path(output_dir, "02-gex_clustered.qs"))
```

------------------------------------------------------------------------

# Section 5: Assessing Confounders and Deciding on Cell Cycle Regression

## 5.1 Diagnostic DimPlots

Before annotating clusters, we need to understand what is driving their
structure. In a multi-donor experiment, ideal clusters should reflect
**cell type or cell state** — not donor of origin, sequencing run,
treatment condition, or cell cycle phase. Each of these would represent
a confounder that could lead to misannotation or incorrect biological
conclusions.

We generate a systematic set of DimPlots, coloring by different metadata
variables, to evaluate potential sources of confounding.

```{r diagnostic-dimplots, fig.width=14, fig.height=10}
# A color palette with one distinct color per donor (38 donors)
donor_palette <- c(
  "#e1ab33","#6572e7","#75b235","#9943bf","#53c35e","#bc73e9","#b6be38","#6751b6",
  "#478734","#dc60c5","#5dbf8c","#d33b88","#40c0bc","#d94b2d","#5e91e3","#dc8331",
  "#435ea7","#a88424","#93429c","#939738","#af79cd","#9db46c","#a4437a","#36815b",
  "#d24258","#4fa7d6","#ae502a","#787dbc","#d1ab67","#835a95","#636925","#bf9fe1",
  "#966933","#da75b0","#e58f6c","#924861","#e28aa6","#ba6262"
)

# Treatment — if cells from different conditions cluster separately, treatment
# is driving cluster structure before we even look at cell type. Some
# separation is expected (lead does change gene expression), but complete
# segregation by condition at this step would be a red flag.
p1 <- DimPlot(gex_subset, group.by = "treatment", pt.size = 0.3,
              shuffle = TRUE, seed = 42) +
  labs(title = "Treatment condition")

# Sequencing run — cells from the same run should be interleaved with other
# runs. Distinct run-specific clusters indicate a technical batch effect.
p2 <- DimPlot(gex_subset, group.by = "orig.ident", pt.size = 0.3,
              shuffle = TRUE, seed = 42) +
  labs(title = "Sequencing batch")

# Cell cycle phase — do cells cluster by phase? If so, regression may help.
p3 <- DimPlot(gex_subset, group.by = "Phase", pt.size = 0.3,
              shuffle = TRUE, seed = 42) +
  labs(title = "Cell cycle phase")

# Donor — the gold standard check for genetic batch effects. You want to see
# all 40 donor colors interleaved throughout every cluster. Distinct donor
# "islands" would indicate that some donor cell lines may have differentiated and should be followed up on more cosely.
p4 <- DimPlot(gex_subset, group.by = "donorlabel", pt.size = 0.3,
              shuffle = TRUE, seed = 42) +
  scale_color_manual("Donor", values = donor_palette) +
  NoLegend() +
  labs(title = "Donor")

(p1 + p2) / (p3 + p4)

ggsave(here("Analysis_Outputs", "5-1_confounder_dimplots.png"),
       width = 14, height = 10, dpi = 300)

```

```{r faceted-by-treatment, fig.width=14, fig.height=4}
# Faceting by treatment makes it easier to see whether one condition is
# pulling cells into a distinct region of the UMAP.
DimPlot(gex_subset, group.by = "treatment", pt.size = 0.3,
        split.by = "treatment") + NoLegend()

ggsave(here("Analysis_Outputs", "5-1_umap_faceted_by_treatment.png"),
       width = 14, height = 4, dpi = 300)
```

```{r plot-donors, fig.width=14, fig.height=4}
DimPlot(gex_subset, group.by = "donor", pt.size = 0.3,
        cols = donor_palette,
        shuffle = TRUE, seed = 42)

ggsave(here("Analysis_Outputs", "5-1_umap_by_donor.png"),
       width = 14, height = 6, dpi = 300)
```

## 5.2 Cell Cycle Regression — To Regress or Not

Cell cycle regression is a common step in NPC analyses because these
cells are actively proliferating and cell cycle phase can dominate gene
expression variation. However, regression is not always necessary and
carries a cost: removing cell cycle signal entirely can obscure real
biology if you are interested in proliferation itself.

A conservative approach is to regress out only the **difference**
between S and G2M scores (`CC.Difference = S.Score - G2M.Score`). This
removes the within-cycling variation (whether a cycling cell is in S vs
G2M) while preserving the distinction between cycling (S or G2M) and
non-cycling (G1) cells.

Look at the Phase DimPlot you generated above. If cycling cells form a
distinct cluster separated from non-cycling cells of the same type,
regression is likely warranted. If phase is broadly distributed across
clusters, the evidence for regression is weaker.

```{r cell-cycle-regression}
# Compute the CC.Difference score — this is what we regress out
gex_subset$CC.Difference <- gex_subset$S.Score - gex_subset$G2M.Score

# Re-scale regressing out CC.Difference, then re-run PCA, clustering, UMAP
gex_CCR <- ScaleData(gex_subset, vars.to.regress = "CC.Difference")
gex_CCR <- RunPCA(gex_CCR, npcs = pca_npcs)
gex_CCR <- gex_CCR %>%
  FindNeighbors(dims = 1:npcs) %>%
  FindClusters(resolution = 0.2) %>%
  RunUMAP(dims = 1:npcs)
```

```{r compare-ccr-dimplots, fig.width=14, fig.height=5}
# Compare Phase structure before and after regression.
# After regression, cell cycle phase should be distributed evenly across
# all clusters rather than forming phase-specific clusters.
p_before <- DimPlot(gex_subset, group.by = "Phase", pt.size = 0.3) +
  labs(title = "Before CCR")

p_after  <- DimPlot(gex_CCR,    group.by = "Phase", pt.size = 0.3) +
  labs(title = "After CCR")

p_before + p_after

ggsave(here("Analysis_Outputs", "5-2_ccr_before_after.png"),
       width = 14, height = 5, dpi = 300)
```

```{r save-both-objects}
# Save both versions. Choose which to load at the start of Section 6
# based on what you observed in the diagnostic plots.
qsave(gex_subset, file.path(output_dir, "03-gex_no_CCR.qs"))
qsave(gex_CCR,    file.path(output_dir, "03-gex_CCR.qs"))

message("Both objects saved. Load the appropriate one at the start of Section 6.")
```

Here, we will save both Seurat objects (with and without CC regression)
so you can decide which to use for downstream steps including
annotation.

------------------------------------------------------------------------

# Section 6: Cell Type Annotation

## 6.1 From Cluster Numbers to Cell Types

Unsupervised clustering gives you numbered groups. Those numbers are
analytically useful but biologically meaningless — they tell you that
cells in cluster 3 are more similar to each other than to cells in
cluster 7, but not what either group actually is. Annotation is the
process of assigning biological identity to each cluster based on the
genes it expresses.

The standard approach is to identify **marker genes**: genes that are
preferentially expressed in one cluster relative to all others. We use
`presto::wilcoxauc()`, an extremely fast implementation of the Wilcoxon
rank-sum test, which returns an AUC (area under the ROC curve) for every
gene in every cluster. An AUC of 1 means the gene perfectly separates
that cluster from all others; an AUC of 0.5 means it does no better than
chance. We rank markers by AUC.

## 6.2 Load and Inspect the Clustered Object

```{r load-for-annotation, message=FALSE}
library(presto)  # Fast Wilcoxon-based marker testing

# Load whichever object you decided to use at the end of Section 5.
# Comment out the one you are not using.
#gex <- qread(file.path(output_dir, "03-gex_no_CCR.qs"))   # no cell cycle regression
# gex <- qread(file.path(output_dir, "03-gex_CCR.qs"))    # with cell cycle regression

gex <- gex_subset
rm(gex_subset)
gc()

Idents(gex) <- "seurat_clusters"

# Confirm the cluster structure and plot with a nice plotting package
DimPlot_scCustom(gex, 
                 shuffle = TRUE, 
                 label = TRUE, 
                 label.box = TRUE, 
                 label.size = 4, 
                 label.color = "black",
                 repel = TRUE) + 
  NoLegend() + 
  labs(title = "Seurat Clusters")

ggsave(here("Analysis_Outputs", "6-2_umap_seurat_clusters.png"),
       width = 8, height = 7, dpi = 300,  bg = "white")

table(gex$seurat_clusters)
```

## 6.3 Finding Marker Genes

```{r find-markers}
# wilcoxauc() tests every gene against every cluster in one call.
# group_by = "seurat_clusters" tells it which grouping variable to use.
# assay = "data" uses the log-normalized counts (not raw).
cluster_markers <- wilcoxauc(
  gex,
  group_by     = "seurat_clusters",
  assay        = "data",
  seurat_assay = "RNA"
)

# Set factor levels for ordered output
n_clusters <- length(unique(gex$seurat_clusters))
cluster_markers <- cluster_markers %>%
  mutate(group = factor(group, levels = as.character(0:(n_clusters - 1))))

# Extract top 10 markers per cluster ranked by AUC
top_markers_per_cluster <- Extract_Top_Markers(
  marker_dataframe = cluster_markers,
  num_genes        = 10,
  group_by         = "group",
  rank_by          = "auc",
  gene_column      = "feature"
)

# Print the top markers — note these are Ensembl IDs;
# we will relabel with gene symbols for the dot plot below
head(top_markers_per_cluster, 20)
```

```{r convert-markers-to-symbols}
# Convert the top marker Ensembl IDs to gene symbols for readability
n_top_genes <- min(10 * n_clusters, length(unique(top_markers_per_cluster)))

top_marker_symbols <- as_tibble(unique(top_markers_per_cluster[seq_len(n_top_genes)])) %>%
  inner_join(ensg_to_symbol, by = c("value" = "gene_id")) %>%
  distinct(value, .keep_all = TRUE)

head(top_marker_symbols)
```

## 6.4 Visualizing Marker Expression

```{r dotplot-markers, fig.width=12, fig.height=20}
# DotPlot encodes two pieces of information per gene per cluster:
# - Dot COLOR: mean normalized expression in that cluster
# - Dot SIZE: fraction of cells in that cluster expressing the gene
# Together these let you distinguish a marker that is highly expressed
# in many cells (ideal) from one that is highly expressed in only a few.
DotPlot_scCustom(
  seurat_object = gex,
  features      = unique(top_markers_per_cluster[seq_len(n_top_genes)]),
  colors_use    = viridis_plasma_dark_high,
  flip_axes     = TRUE,
  x_lab_rotate  = TRUE
) + scale_x_discrete(labels = top_marker_symbols$gene_name) +
  labs(title = "Top 10 Marker Genes per Cluster (by AUC)")

ggsave(here("Analysis_Outputs", "6-4_dotplot_top_markers_per_cluster.png"),
       width = 12, height = 20, dpi = 300,  bg = "white")
```

```{r npc-marker-dotplot, fig.width=10, fig.height=8}
# The NPC marker gene lists provide curated reference markers for
# distinguishing NPCs from neurons, neural crest, and other populations.
# Use these alongside the data-driven markers above.
NPCvsOther  <- read_csv(here("NPCvsOther_genelist.csv"))  %>% arrange(celltype_or_region)
NPCshort    <- read_csv(here("NPCmarkers_shortgenelist.csv")) %>% arrange(celltype_or_region)

# Broad NPC vs other cell type markers
DotPlot_scCustom(
  seurat_object = gex,
  features      = NPCvsOther$gene_id,
  colors_use    = viridis_plasma_dark_high,
  flip_axes     = TRUE
) + scale_x_discrete(labels = NPCvsOther$gene_name) +
  labs(title = "NPC vs Other Cell Type Markers")

ggsave(here("Analysis_Outputs", "6-4_dotplot_npc_vs_other_markers.png"),
       width = 10, height = 8, dpi = 300,  bg = "white")
```

```{r npc-shortlist-dotplot, fig.width=8, fig.height=6}
# Shortlist of canonical NPC markers
DotPlot_scCustom(
  seurat_object = gex,
  features      = NPCshort$gene_id,
  colors_use    = viridis_plasma_dark_high,
  flip_axes     = TRUE
) + scale_x_discrete(labels = NPCshort$gene_name) +
  labs(title = "Canonical NPC Marker Shortlist")

ggsave(here("Analysis_Outputs", "6-4_dotplot_npc_shortlist_markers.png"),
       width = 8, height = 6, dpi = 300,  bg = "white")
```

```{r feature-plots-npc, fig.width=14, fig.height=10}
# FeaturePlots show where individual marker genes are expressed on the UMAP.
# Use these alongside the dot plot to build confidence in your annotations.
# Choose 4-6 genes that together discriminate your expected cell types.
plots_by_type <- list()

for (ct in unique(NPCshort$celltype_or_region)) {
  
  genes   <- NPCshort$gene_id[NPCshort$celltype_or_region == ct]
  symbols <- NPCshort$gene_name[NPCshort$celltype_or_region == ct]
  
  p <- FeaturePlot_scCustom(
    gex,
    features    = genes,
    reduction   = "umap",
    order       = TRUE,
    num_columns = 3
  )
  
  # Overwrite titles with gene symbols
  for (i in seq_along(genes)) {
    p[[i]] <- p[[i]] + ggtitle(symbols[i])
  }
  
  # Add celltype_or_region as patchwork title
  plots_by_type[[ct]] <- p + plot_annotation(title = ct)
}

# Print all
for (ct in names(plots_by_type)) {
  ggsave(here("Analysis_Outputs", paste0("6-4_featureplot_", ct, ".png")),
       plot   = plots_by_type[[ct]],
       width  = 3 * 4,
       height = ceiling(length(NPCshort$gene_id[NPCshort$celltype_or_region == ct]) / 3) * 4,
       dpi    = 300)
print(plots_by_type[[ct]])
}
```

## 6.5 Assigning Cell Type Labels

Now use the marker plots above to assign a biological identity to each
cluster. Look up unfamiliar genes in [CellMarker
2.0](http://bio-bigdata.hrbmu.edu.cn/CellMarker/) or
[PanglaoDB](https://panglaodb.se/).

In this dataset you should expect to find: **NPCs**, **Neurons**
(mature), **Newborn Neurons**, and **Neural Crest** cells. One cluster
(cluster 3 in the original analysis) had an ambiguous marker profile —
keep an eye out for this and consider how you would handle it.

`scCustomize` provides a workflow for recording your annotations in a
CSV file and applying them back to the Seurat object. Run
`Create_Cluster_Annotation_File()` once to generate the template, fill
it in with your cluster labels, then load it back with
`Pull_Cluster_Annotation()`.

```{r create-annotation-template, eval=FALSE}
# Run this ONCE to generate the annotation template CSV.
# Fill in the new_cluster_idents and sub_type columns, then save the file.
# Do NOT re-run this chunk after filling in the template or it will overwrite
# your work.
# Create_Cluster_Annotation_File(
#   file_path = here(),
#   file_name = "cluster_annotation"
# )
```

```{r apply-annotation}
# After filling in cluster_annotation.csv, run this chunk to apply labels.
# Pull_Cluster_Annotation() reads the CSV and returns a named vector.
annotation_info <- Pull_Cluster_Annotation(
  annotation = here("cluster_annotation.csv")
)

# Rename_Clusters() applies the new labels as Seurat identities and
# stores the numeric cluster IDs in a new metadata column for reference.
Idents(gex) = "seurat_clusters"
gex <- Rename_Clusters(
  seurat_object = gex,
  new_idents    = annotation_info$sub_type,
  meta_col_name = "Res0.2_idents"
)

gex$cell_subtype  <- Idents(gex)
Idents(gex)       <- gex$Res0.2_idents
gex <- Rename_Clusters(
  seurat_object = gex,
  new_idents    = annotation_info$new_cluster_idents
)

gex$annotatedclust <- Idents(gex)

# Confirm the assignment
table(Idents(gex))
```

```{r plot-annotated-umap, fig.width=14, fig.height=5}
# Full annotated UMAP
dim_palette <- c("#9b78ff", "cyan", "#f78215", "#005ac7",
                 "#c80053", "#72cd3b", "#006e36", "#a54600",
                 "#763bc0", "#ff77f1")

p_full <- DimPlot(gex, cols = dim_palette, label = TRUE,
                  label.size = 3, repel = TRUE) +
  labs(title = "Annotated Cell Types")

# Highlight each population individually
p_npc  <- Cluster_Highlight_Plot(gex, cluster_name = "NPCs",
                                  highlight_color = "#9b78ff",
                                  background_color = "lightgray")
p_neu  <- Cluster_Highlight_Plot(gex, cluster_name = "Neurons",
                                  highlight_color = "cyan",
                                  background_color = "lightgray")
p_nc   <- Cluster_Highlight_Plot(gex, cluster_name = "Neural Crest",
                                  highlight_color = "#f78215",
                                  background_color = "lightgray")
#p_nbn  <- Cluster_Highlight_Plot(gex, cluster_name = "Newborn Neurons",
#                                  highlight_color = "#005ac7",
#                                  background_color = "lightgray")

p_full / (p_npc + p_neu + p_nc)
ggsave(here("Analysis_Outputs", "6-5_umap_annotated_clusters.png"),
       plot   = p_full / (p_npc + p_neu + p_nc),
       width  = 14,
       height = 16,
       dpi    = 300,
       bg     = "white")
```

```{r save-annotated-object}
# Add a group_donor column — donor x treatment combination used in pseudobulk
gex$group_donor <- paste0(gex$treatment, "_", gex$donor)

qsave(gex, file.path(output_dir, "04-gex_annotated.qs"))
```

------------------------------------------------------------------------

# Section 7: Pseudobulk Differential Expression with dreamlet

## 7.1 Why Pseudobulking

A common mistake in scRNA-seq differential expression is treating each
cell as an independent observation. With 40 donors each contributing
hundreds or thousands of cells, a naive test comparing Control vs Low Pb
cells would have tens of thousands of "samples" per condition —
massively inflating statistical power and producing enormous numbers of
false positives.

The correct unit of biological replication in this experiment is the
**donor**, not the cell. Ten cells from the same donor are not ten
independent observations — they share a genome, an epigenome, and a
culture history.

**Pseudobulking** solves this by aggregating (summing) raw counts across
all cells from the same donor and cell type, producing one pseudo-sample
per donor per cell type. We then run a bulk RNA-seq style mixed model on
these aggregated counts. The `dreamlet` package implements this workflow
with a voom-style normalization followed by a linear mixed model that
explicitly models donor as a random effect.

Our model formula will be:

```         
~ (1|donor) + treatment + sex + PropPhase_G2M + PropPhase_G1
```

-   `(1|donor)` — donor is a **random effect**. This accounts for the
    non-independence of cells from the same donor and lets the model
    generalize across the 40-donor population rather than fitting
    donor-specific intercepts. With 38 donors, fitting donor as a fixed
    effect would consume 37 degrees of freedom and likely overfit.
-   `treatment` — the **fixed effect of interest**. Coefficients for
    `treatmentLow` and `treatmentHigh` give us the effect of each dose
    relative to Control.
-   `sex` — a fixed covariate to control for sex differences in baseline
    expression.
-   `PropPhase_G2M`, `PropPhase_G1` — the proportion of cells from each
    donor-treatment group in each cell cycle phase, included to absorb
    any residual cell cycle variation not addressed in Section 5.

## 7.2 Preparing the Pseudobulk Object

```{r prepare-pseudobulk, message=FALSE}
library(SingleCellExperiment)  # Bioconductor class required by dreamlet
library(dreamlet)              # Pseudobulk DE framework
library(variancePartition)     # Mixed-model variance decomposition
library(BiocParallel)          # Parallel processing backend
library(qvalue)                # pi1 statistic for signal estimation
library(ggcorrplot)            # Covariate correlation heatmap

# Load the annotated object
gex <- qread(file.path(output_dir, "04-gex_annotated.qs"))

# Subset to NPCs — the cell type of primary interest for lead neurotoxicity
gex_NPC <- subset(gex, idents = "NPCs")
rm(gex); gc()

message(sprintf("NPCs retained: %d cells", ncol(gex_NPC)))
table(gex_NPC$treatment)
```

```{r compute-phase-proportions}
# Compute the proportion of cells in each cell cycle phase per
# donor-treatment group. These will be used as covariates in the model
# to absorb any residual cell cycle variation.
clusters_phase <- gex_NPC[[]] %>%
  group_by(group_donor, Res0.2_idents) %>%
  summarize(
    PropPhase_G1  = sum(Phase == "G1")  / n(),
    PropPhase_G2M = sum(Phase == "G2M") / n(),
    PropPhase_S   = sum(Phase == "S")   / n(),
    .groups = "drop"
  )

# Remove any existing PropPhase columns and add the freshly computed ones
gex_NPC@meta.data <- gex_NPC@meta.data %>%
  select(-starts_with("PropPhase"))

gex_NPC@meta.data <- gex_NPC@meta.data %>%
  bind_cols(
    left_join(gex_NPC[[]], clusters_phase,
              by = c("Res0.2_idents", "group_donor")) %>%
      select(starts_with("PropPhase"))
  )
```

```{r convert-to-sce-and-pseudobulk}
# Convert to SingleCellExperiment — required by dreamlet
gex_NPC.sce <- as.SingleCellExperiment(gex_NPC)
rm(gex_NPC); gc()

# Set factor levels so Control is the reference group in the model
gex_NPC.sce$treatment <- factor(gex_NPC.sce$treatment,
                                 levels = c("Control", "Low", "High"))

# aggregateToPseudoBulk() sums raw counts per donor per cell type.
# cluster_id: which metadata column identifies cell types
# sample_id:  the unit of replication (donor x treatment)
# Donors with fewer than the minimum number of cells in a cell type
# will be automatically excluded from that cell type's model — this is
# expected given the uneven donor distribution we noted in Section 2.
pb <- aggregateToPseudoBulk(
  gex_NPC.sce,
  assay       = "counts",
  cluster_id  = "ident",
  sample_id   = "group_donor",
  checkValues = TRUE,
  verbose     = TRUE,
  BPPARAM     = SnowParam(4, progressbar = TRUE)
)

# How many pseudo-samples do we have per cell type?
table(pb$treatment)
```

## 7.3 Model Specification and Voom Normalization

```{r check-covariate-correlations, fig.width=5, fig.height=4}
# Before modeling, check whether covariates are collinear.
# canCorPairs() computes canonical correlations between model terms.
# Values close to 1 indicate redundant covariates that would destabilize
# the model. Treatment and sex should be essentially uncorrelated in a
# well-designed experiment.
form_check <- ~ treatment + sex
C <- canCorPairs(form_check, colData(pb))
ggcorrplot(C, hc.order = TRUE, lab = TRUE) +
  labs(title = "Covariate canonical correlations")
ggsave(here("Analysis_Outputs", "6-5_donor_cluster_composition.png"),
       width  = 10,
       height = 8,
       dpi    = 300,
       bg     = "white")
```

```{r voom-normalization, fig.width=10, fig.height=4}
# Define the full model formula
form_dreamlet <- ~ (1|donor) + treatment + sex + PropPhase_G2M + PropPhase_G1

# processAssays() performs TMM normalization and voom transformation.
# Voom estimates the mean-variance relationship in count data and
# produces precision weights for each observation. A well-behaved voom
# plot shows a decreasing trend from left to right — high-count genes
# have lower variability relative to their mean.
res.proc <- processAssays(
  pb,
  form_dreamlet,
  BPPARAM = SnowParam(4, progressbar = TRUE)
)

plotVoom(res.proc, ncol = 4)

qsave(res.proc, file.path(output_dir, "05-res_proc.qs"))
```

## 7.4 Running dreamlet

```{r run-dreamlet}
# dreamlet() fits the linear mixed model for every gene in every cell type.
# This is the main DE step — it may take a few minutes.
res.dl <- dreamlet(
  res.proc,
  form_dreamlet,
  BPPARAM = SnowParam(4, progressbar = TRUE)
)

qsave(res.dl, file.path(output_dir, "06-res_dl.qs"))

# What coefficients are available? These correspond to the model terms.
# treatmentLow and treatmentHigh are the contrasts of interest.
coefNames(res.dl)
```

```{r extract-deg-tables}
# Extract the full results table for each contrast.
# topTable() returns all genes sorted by p-value, with logFC and
# Benjamini-Hochberg adjusted p-values.
low_effect_genes <- variancePartition::topTable(
  res.dl[["NPCs"]],
  coef     = "treatmentLow",
  adjust   = "BH",
  number   = Inf,
  sort.by  = "P"
) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("gene_id") %>%
  left_join(ensg_to_symbol %>% select(gene_id, gene_name), by = "gene_id")

high_effect_genes <- variancePartition::topTable(
  res.dl[["NPCs"]],
  coef     = "treatmentHigh",
  adjust   = "BH",
  number   = Inf,
  sort.by  = "P"
) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("gene_id") %>%
  left_join(ensg_to_symbol %>% select(gene_id, gene_name), by = "gene_id")

# How many DEGs at FDR < 0.05?
message(sprintf("DEGs (FDR < 0.05) — Low Pb:  %d", sum(low_effect_genes$adj.P.Val  < 0.05)))
message(sprintf("DEGs (FDR < 0.05) — High Pb: %d", sum(high_effect_genes$adj.P.Val < 0.05)))

# Summarize signal across all assays (cell types)
Low_stats <- topTable(res.dl, coef = "treatmentLow", number = Inf) %>%
  as_tibble() %>%
  group_by(assay) %>%
  summarize(
    nDE    = sum(adj.P.Val < 0.05),
    pi1    = 1 - qvalue::pi0est(P.Value)$pi0,
    nGenes = n()
  )

print(Low_stats)
```

## 7.5 Variance Partition

Variance partition answers a different question from differential
expression: rather than asking *whether* treatment changes expression of
a given gene, it asks *how much* of that gene's total expression
variance is attributable to each factor in the model — donor, treatment,
sex, or residual noise.

This is a powerful diagnostic. If donor explains the vast majority of
variance for most genes, it confirms that genetic background is a
dominant driver of NPC transcription, and underscores why the mixed
model design is essential.

```{r variance-partition, fig.width=14, fig.height=5}
# fitVarPart() fits the variance partition model and returns per-gene
# variance fractions for each model term.
varpart_formula <- ~ (1|donor) + (1|treatment) + (1|sex) +
                     PropPhase_G2M + PropPhase_G1

vp <- fitVarPart(
  res.proc,
  varpart_formula,
  BPPARAM = SnowParam(4, progressbar = TRUE)
)

# sortCols() orders the columns by median variance explained across genes
plotVarPart(sortCols(vp), label.angle = 60, ncol = 4) +
  labs(title = "Variance Partition — NPC gene expression")
ggsave(here("Analysis_Outputs", "7-5_variance_partition.png"),
       width = 14, height = 5, dpi = 300, bg = "white")

qsave(vp, file.path(output_dir, "07-variance_partition.qs"))
```

------------------------------------------------------------------------

# Section 8: Visualizing Differential Expression

## 8.1 Volcano Plots

A volcano plot places every gene in a two-dimensional space defined by
effect size (log2 fold change, x-axis) and statistical confidence
(−log10 adjusted p-value, y-axis). Genes in the upper-right quadrant are
significantly upregulated; genes in the upper-left quadrant are
significantly downregulated. Genes near the x-axis are not significant
regardless of their fold change; genes near the y-axis are statistically
significant but have small effect sizes.

We use two thresholds to classify genes as "key hits": \|logFC\| \> 0.5
and adjusted p-value \< 0.05. The 0.5 log2 fold change threshold is
somewhat arbitrary — it represents approximately a 1.4-fold change — but
is a common convention for identifying biologically meaningful changes.

```{r volcano-low, fig.width=7, fig.height=7}
# Volcano plot for Low Pb (3 µM) vs Control
low_effect_genes %>%
  mutate(
    direction = case_when(
      logFC >  0.5 & adj.P.Val < 0.05 ~ "UP",
      logFC < -0.5 & adj.P.Val < 0.05 ~ "DOWN",
      TRUE ~ "NS"
    ),
    label = if_else(direction %in% c("UP", "DOWN"), gene_name, NA_character_)
  ) %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = direction, label = label)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "gray40") +
  geom_hline(yintercept = -log10(0.05),  linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c("NS" = "grey70", "UP" = "#276bc4", "DOWN" = "#cc2132")) +
  geom_text_repel(size = 3, max.overlaps = 8, na.rm = TRUE) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = "Low Pb (3 µM) vs Control — NPCs",
       x = "log2 Fold Change", y = expression(-log[10]("adj. p-value")))


ggsave(here("Analysis_Outputs", "8-1_volcano_low_pb.png"),
       width = 7, height = 7, dpi = 300, bg = "white")
```

```{r volcano-high, fig.width=7, fig.height=7}
# Volcano plot for High Pb (10 µM) vs Control
high_effect_genes %>%
  mutate(
    direction = case_when(
      logFC >  0.5 & adj.P.Val < 0.05 ~ "UP",
      logFC < -0.5 & adj.P.Val < 0.05 ~ "DOWN",
      TRUE ~ "NS"
    ),
    label = if_else(direction %in% c("UP", "DOWN"), gene_name, NA_character_)
  ) %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = direction, label = label)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "gray40") +
  geom_hline(yintercept = -log10(0.05),  linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c("NS" = "grey70", "UP" = "#276bc4", "DOWN" = "#cc2132")) +
  geom_text_repel(size = 3, max.overlaps = 8, na.rm = TRUE) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = "High Pb (10 µM) vs Control — NPCs",
       x = "log2 Fold Change", y = expression(-log[10]("adj. p-value")))

ggsave(here("Analysis_Outputs", "8-1_volcano_high_pb.png"),
       width = 7, height = 7, dpi = 300, bg = "white")
```

## 8.2 Pathway-Annotated Volcano

When you have a specific biological hypothesis — for example, that lead
exposure activates oxidative stress pathways — you can color volcano
plot points by pathway membership rather than just by direction. This
makes it immediately visible whether your hypothesis genes are enriched
among the significant hits, and in which direction they move.

```{r pathway-volcano, fig.width=9, fig.height=8}
# Known gene sets from lead neurotoxicity literature
keap_nfe2l2_genes <- c("NQO1","CDKN1A","EGF","SESN1","TXNRD1","ME1",
                        "HMOX1","SLC7A11","GCLM","TKT")
tp53_genes        <- c("FAS","TNFRSF10A","TNFRSF10D")
elev_ca2_genes    <- c("EGF","RARRES2","SERPINE1","CD9","VEGFD","CLU","TTN")

high_effect_genes %>%
  mutate(
    SYMBOL = toupper(gene_name),
    pathway = case_when(
      SYMBOL %in% keap_nfe2l2_genes    ~ "KEAP1-NFE2L2",
      grepl("^MT1|^MT2", SYMBOL)       ~ "Metallothioneins",
      SYMBOL %in% tp53_genes           ~ "TP53 signaling",
      SYMBOL %in% elev_ca2_genes       ~ "Elevated cytosolic Ca2+",
      TRUE                             ~ "Other"
    ),
    label = if_else(pathway != "Other" & adj.P.Val < 0.05, SYMBOL, NA_character_)
  ) %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val),
             color = pathway, label = label)) +
  geom_point(alpha = 0.8, size = 1.5) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "gray40") +
  geom_hline(yintercept = -log10(0.05),  linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c(
    "Other"                 = "grey80",
    "KEAP1-NFE2L2"          = "#cc2132",
    "Metallothioneins"      = "#1e9d89",
    "TP53 signaling"        = "#b26fd6",
    "Elevated cytosolic Ca2+" = "#f39c12"
  )) +
  geom_text_repel(size = 3.5, max.overlaps = 20, na.rm = TRUE,
                  fontface = "bold") +
  theme_classic() +
  labs(color  = "Pathway",
       title  = "High Pb (10 µM) — pathway-annotated volcano",
       x      = "log2 Fold Change",
       y      = expression(-log[10]("adj. p-value")))


ggsave(here("Analysis_Outputs", "8-2_volcano_high_pb_pathway_annotated.png"),
       width = 9, height = 8, dpi = 300, bg = "white")
```

## 8.3 Overlap Between Conditions

How much do the transcriptional responses to Low and High Pb share in
common? Genes shared between both conditions are likely core responses
to Pb regardless of dose. Condition-specific genes may reflect
dose-dependent mechanisms or off-target effects of the higher
concentration.

```{r venn-diagram, fig.width=6, fig.height=5, message=FALSE}
library(ggVennDiagram)

low_sig  <- low_effect_genes  %>% filter(abs(logFC) >= 0.5 & adj.P.Val <= 0.05) %>% pull(gene_name) %>% unique() %>% na.omit()
high_sig <- high_effect_genes %>% filter(abs(logFC) >= 0.5 & adj.P.Val <= 0.05) %>% pull(gene_name) %>% unique() %>% na.omit()

ggVennDiagram(list(Low = low_sig, High = high_sig), label_alpha = 0) +
  scale_fill_gradient(low = "#FCE2E1", high = "#F06060") +
  labs(title = "Overlap of significant DEGs: Low vs High Pb (|logFC| > 0.5, FDR < 0.05)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")

ggsave(here("Analysis_Outputs", "8-3_venn_deg_overlap_low_vs_high.png"),
       width = 6, height = 5, dpi = 300, bg = "white")
```

```{r save-deg-tables}
# Save DEG tables — used in Section 9 for GSEA
saveRDS(low_effect_genes,  file.path(output_dir, "08-DEGs_Low.rds"))
saveRDS(high_effect_genes, file.path(output_dir, "08-DEGs_High.rds"))
```

------------------------------------------------------------------------

# Section 9: Gene Set Enrichment Analysis with fgsea

## 9.1 From Gene Lists to Pathways

A DEG list tells you which individual genes change — but biological
processes are not enacted by single genes, they emerge from coordinated
changes across many genes working in the same pathway. Gene Set
Enrichment Analysis (GSEA) shifts the question from "which genes are
significant?" to "which pathways are systematically enriched among the
most strongly changed genes?"

GSEA works on a **ranked gene list** rather than a binary
significant/not significant classification. Every gene in the experiment
is ranked by a score that encodes both direction and significance — here
we use −log10(adj.P.Val) × sign(logFC), so genes at the top of the list
are significantly upregulated and genes at the bottom are significantly
downregulated. GSEA then asks: for a given gene set (e.g., all genes in
the Reactome "DNA Repair" pathway), are members of that set concentrated
at the top or bottom of the ranked list, or scattered randomly
throughout?

The output is a **Normalized Enrichment Score (NES)**: positive NES
means the pathway is enriched among upregulated genes; negative NES
means it is enriched among downregulated genes. P-values are estimated
by permutation.

An important advantage of this approach over simple overlap tests
(over-representation analysis) is that it uses the full ranked list —
genes that don't pass a significance threshold but consistently move in
the same direction still contribute to pathway enrichment.

## 9.2 Loading Pathway Gene Sets

```{r load-pathways, message=FALSE}
library(fgsea)     # Fast GSEA implementation (Korotkevich et al.)
library(msigdbr)   # Programmatic access to MSigDB

# We query the Molecular Signatures Database (MSigDB) for curated pathway
# gene sets. The gene sets use Ensembl IDs, which match our DEG tables.

# C2: Curated gene sets — Reactome, BioCarta, WikiPathways
reactome <- msigdbr(species = "Homo sapiens",
                    category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(reactome$ensembl_gene, reactome$gs_name)

biocarta <- msigdbr(species = "Homo sapiens",
                    category = "C2", subcategory = "CP:BIOCARTA")
biocarta.list <- split(biocarta$ensembl_gene, biocarta$gs_name)

wiki <- msigdbr(species = "Homo sapiens",
                category = "C2", subcategory = "CP:WIKIPATHWAYS")
wiki.list <- split(wiki$ensembl_gene, wiki$gs_name)

# Combine curated sets into one searchable list
c2_list <- c(reactome.list, biocarta.list, wiki.list)

# C5: Gene Ontology biological process, molecular function, cellular component
go <- msigdbr(species = "Homo sapiens", category = "C5")
go.list <- split(go$ensembl_gene, go$gs_name)

message(sprintf("Reactome: %d gene sets", length(reactome.list)))
message(sprintf("Combined C2: %d gene sets", length(c2_list)))
message(sprintf("GO (C5): %d gene sets",     length(go.list)))
```

## 9.3 Running fgsea

```{r run-fgsea}
# Helper function to build the ranked gene list and run fgseaMultilevel
run_gsea <- function(deg_table, pathway_list, seed = 19) {
  # Rank genes by signed significance
  # Genes at the top: significantly upregulated
  # Genes at the bottom: significantly downregulated
  ranked <- -log10(deg_table$adj.P.Val) * sign(deg_table$logFC)
  names(ranked) <- deg_table$gene_id
  ranked <- sort(ranked, decreasing = TRUE)

  set.seed(seed)
  fgseaMultilevel(
    pathways = pathway_list,
    stats    = ranked,
    minSize  = 15,    # Pathways smaller than this are too noisy
    maxSize  = 500    # Pathways larger than this are too broad to interpret
  )
}

# Run against Reactome for both conditions
gsea_low_reactome  <- run_gsea(low_effect_genes,  reactome.list)
gsea_high_reactome <- run_gsea(high_effect_genes, reactome.list)

# Top significant pathways for Low Pb
gsea_low_reactome %>%
  filter(padj < 0.05) %>%
  arrange(padj) %>%
  select(pathway, NES, padj, size) %>%
  head(20)
```

## 9.4 Visualizing GSEA Results

```{r plot-gsea-low, fig.width=11, fig.height=7}
# Lollipop plot of top enriched and depleted Reactome pathways — Low Pb
top_up   <- gsea_low_reactome %>% filter(NES > 0) %>% slice_min(padj, n = 10)
top_down <- gsea_low_reactome %>% filter(NES < 0) %>% slice_min(padj, n = 10)
top_low  <- bind_rows(top_up, top_down)

ggplot(top_low, aes(x = reorder(pathway, NES), y = NES)) +
  geom_segment(aes(xend = reorder(pathway, NES), yend = 0), color = "gray60") +
  geom_point(aes(fill = ifelse(padj < 0.05, NES > 0, "ns")),
             pch = 21, size = 4, stroke = 0.8) +
  scale_fill_manual(
    name   = "Direction",
    values = c("TRUE" = "#cc2132", "FALSE" = "#276bc4", "ns" = "white"),
    labels = c("TRUE" = "Upregulated", "FALSE" = "Downregulated",
               "ns"   = "Not significant")
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  ylim(c(-3, 3)) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(x = NULL, y = "Normalized Enrichment Score (NES)",
       title = "Top Reactome Pathways — Low Pb (3 µM) vs Control")

ggsave(here("Analysis_Outputs", "9-4_gsea_reactome_low_pb.png"),
       width = 11, height = 7, dpi = 300, bg = "white")
```

```{r plot-gsea-high, fig.width=11, fig.height=7}
# Lollipop plot for High Pb
top_up    <- gsea_high_reactome %>% filter(NES > 0) %>% slice_min(padj, n = 10)
top_down  <- gsea_high_reactome %>% filter(NES < 0) %>% slice_min(padj, n = 10)
top_high  <- bind_rows(top_up, top_down)

ggplot(top_high, aes(x = reorder(pathway, NES), y = NES)) +
  geom_segment(aes(xend = reorder(pathway, NES), yend = 0), color = "gray60") +
  geom_point(aes(fill = ifelse(padj < 0.05, NES > 0, "ns")),
             pch = 21, size = 4, stroke = 0.8) +
  scale_fill_manual(
    name   = "Direction",
    values = c("TRUE" = "#cc2132", "FALSE" = "#276bc4", "ns" = "white"),
    labels = c("TRUE" = "Upregulated", "FALSE" = "Downregulated",
               "ns"   = "Not significant")
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  ylim(c(-3, 3)) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(x = NULL, y = "Normalized Enrichment Score (NES)",
       title = "Top Reactome Pathways — High Pb (10 µM) vs Control")

ggsave(here("Analysis_Outputs", "9-4_gsea_reactome_high_pb.png"),
       width = 11, height = 7, dpi = 300, bg = "white")

```

```{r compare-shared-pathways, fig.width=9, fig.height=7}
# Compare NES values for the same pathways across Low and High Pb.
# Pathways that appear in both conditions with consistent NES direction
# represent robust, dose-independent lead responses. Pathways where NES
# changes direction or magnitude between doses reveal dose-dependent biology.

shared_pathways <- inner_join(
  gsea_low_reactome  %>% select(pathway, NES_low  = NES, padj_low  = padj),
  gsea_high_reactome %>% select(pathway, NES_high = NES, padj_high = padj),
  by = "pathway"
) %>%
  filter(padj_low < 0.05 | padj_high < 0.05)

ggplot(shared_pathways, aes(x = NES_low, y = NES_high)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "gray60", linewidth = 0.4) +
  geom_text_repel(
    data = shared_pathways %>% filter(abs(NES_low) > 1.5 | abs(NES_high) > 1.5),
    aes(label = pathway), size = 2.5, max.overlaps = 10
  ) +
  theme_classic() +
  labs(title = "NES comparison: Low vs High Pb (shared significant pathways)",
       x = "NES — Low Pb (3 µM)",
       y = "NES — High Pb (10 µM)")

ggsave(here("Analysis_Outputs", "9-4_gsea_nes_comparison_low_vs_high.png"),
       width = 9, height = 7, dpi = 300, bg = "white")
```

------------------------------------------------------------------------

------------------------------------------------------------------------

# Section 10: WGCNA Co-expression Module-Trait Analysis

## 10.1 Overview

Differential expression tells us which individual genes respond to lead.
WGCNA (Weighted Gene Co-expression Network Analysis) asks a complemen-
tary question: are there **modules** of co-expressed genes whose
coordinated activity correlates with a donor's intrinsic sensitivity or
resistance to lead?

To quantify lead sensitivity per donor, we use **townlet estimates** —
donor-level effect size estimates (β) computed from the same pseudobulk
model. A large positive β means a donor's NPCs showed a strong
transcriptional response to lead; a negative β means their NPCs were
relatively resistant. We correlate each WGCNA module's eigengene (the
first principal component of module expression) with these per-donor
sensitivity scores to identify modules that track with lead response.

## 10.2 Prepare Expression Matrix
```{r wgcna-prep, message=FALSE}
library(WGCNA)

options(stringsAsFactors = FALSE)
enableWGCNAThreads()

# Extract normalized expression matrix from dreamlet processed object
expr_df <- extractData(res.proc, "NPCs")

# Separate gene expression columns (Ensembl IDs) from metadata
gene_cols    <- grepl("^ENSG", colnames(expr_df))
expr_mat_t   <- as.matrix(expr_df[, gene_cols])
rownames(expr_mat_t) <- expr_df$Row.names

message(sprintf("Expression matrix: %d samples x %d genes",
                nrow(expr_mat_t), ncol(expr_mat_t)))

# Check for genes/samples with too many missing values or zero variance
gsg <- goodSamplesGenes(expr_mat_t, verbose = 3)

if (!gsg$allOK) {
  if (sum(!gsg$goodGenes) > 0)
    message(sprintf("Removing %d genes flagged by goodSamplesGenes",
                    sum(!gsg$goodGenes)))
  if (sum(!gsg$goodSamples) > 0)
    message(sprintf("Removing %d samples flagged by goodSamplesGenes",
                    sum(!gsg$goodSamples)))
  expr_mat_t <- expr_mat_t[gsg$goodSamples, gsg$goodGenes]
}

message(sprintf("Final matrix for WGCNA: %d samples x %d genes",
                nrow(expr_mat_t), ncol(expr_mat_t)))

```


```{r plot-sample-tree, fig.width=12, fig.height=5}
par(cex = 0.6, mar = c(0, 4, 2, 0))
plot(sampleTree, main = "Sample clustering to detect outliers",
     sub = "", xlab = "",
     cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)

ggsave <- function(...) invisible(NULL)  # placeholder — use dev.copy for base plots
png(here("Analysis_Outputs", "10-2_wgcna_sample_tree.png"),
    width = 12, height = 5, units = "in", res = 300)
par(cex = 0.6, mar = c(0, 4, 2, 0))
plot(sampleTree, main = "Sample clustering to detect outliers",
     sub = "", xlab = "", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)
dev.off()
```

## 10.3 Load Townlet Sensitivity Estimates
```{r load-townlet}
# Load donor-level lead sensitivity estimates from townlet
townlet <- read_csv(here("townlet_donor_estimates.csv"))

head(townlet)

# Extract donor names from expression matrix rownames
# Format is Treatment_DonorName_Pxx_xxxxxx — strip prefix and suffix
extracted_donors <- sub("^[^_]+_", "", rownames(expr_mat_t))
extracted_donors <- sub("_P[0-9]+_[0-9]+$", "", extracted_donors)

# Keep only samples whose donor is present in townlet
# (Genea16, HUES64, WA7 are excluded as siblings)
keep_samples     <- extracted_donors %in% townlet$donor
expr_mat_t       <- expr_mat_t[keep_samples, ]
extracted_donors <- extracted_donors[keep_samples]

message(sprintf("Samples after removing excluded donors: %d", nrow(expr_mat_t)))
stopifnot(all(extracted_donors %in% townlet$donor))

# Build trait matrix — one row per pseudosample, matched to townlet estimates
# mean_beta: donor's average transcriptional response to lead
# mean_tau:  donor-specific deviation from population mean response
trait_data <- data.frame(
  row.names = rownames(expr_mat_t),
  donor     = extracted_donors
) %>%
  left_join(
    townlet %>% select(donor, mean_beta, mean_tau, sig_vs_baseline),
    by = "donor"
  ) %>%
  mutate(sig_numeric = as.numeric(sig_vs_baseline)) %>%
  select(mean_beta, mean_tau, sig_numeric)

head(trait_data)
```

## 10.4 Choose Soft Thresholding Power
```{r soft-threshold, fig.width=10, fig.height=5}
# WGCNA uses a soft thresholding power β to emphasize strong correlations
# and suppress weak ones, approximating a scale-free network topology.
# We test a range of powers and choose the lowest that achieves a
# scale-free topology fit (R² ≥ 0.85).
powers <- c(1:10, seq(12, 20, by = 2))

sft <- pickSoftThreshold(
  expr_mat_t,
  powerVector  = powers,
  verbose      = 5,
  networkType  = "signed"
)

# Plot scale-free fit and mean connectivity
p1 <- ggplot(data.frame(Power = sft$fitIndices$Power,
                         SFT   = -sign(sft$fitIndices$slope) *
                                   sft$fitIndices$SFT.R.sq),
             aes(x = Power, y = SFT)) +
  geom_point(size = 3) +
  geom_text(aes(label = Power), hjust = -0.3, size = 3) +
  geom_hline(yintercept = 0.85, linetype = "dashed", color = "red") +
  theme_classic() +
  labs(title = "Scale-free topology fit",
       x = "Soft threshold (power)", y = "Scale free topology R²")

p2 <- ggplot(data.frame(Power = sft$fitIndices$Power,
                         MeanK = sft$fitIndices$mean.k.),
             aes(x = Power, y = MeanK)) +
  geom_point(size = 3) +
  geom_text(aes(label = Power), hjust = -0.3, size = 3) +
  theme_classic() +
  labs(title = "Mean connectivity",
       x = "Soft threshold (power)", y = "Mean connectivity")

p1 + p2

ggsave(here("Analysis_Outputs", "10-4_wgcna_soft_threshold.png"),
       width = 10, height = 5, dpi = 300, bg = "white")

# The recommended soft power — look for the knee in the R² plot
# where R² first exceeds 0.85. Adjust this value based on your plot.
soft_power <- sft$powerEstimate
message(sprintf("Recommended soft power: %d", soft_power))
```

## 10.5 Build the Co-expression Network
```{r build-network, message=FALSE}
# blockwiseModules() constructs the weighted gene network, performs
# hierarchical clustering, and assigns genes to modules in one call.
# This is the most compute-intensive step — expect several minutes.
#
# Key parameters:
# power:             soft thresholding power from above
# networkType:       "signed" preserves the direction of correlation
# minModuleSize:     minimum number of genes per module (30 is standard)
# mergeCutHeight:    modules with eigengene correlation > (1 - this value)
#                    are merged; 0.25 is a common default
# TOMType:           "signed" for signed network

soft_power <- 12
net <- blockwiseModules(
  expr_mat_t,
  power             = soft_power,
  networkType       = "signed",
  TOMType           = "signed",
  minModuleSize     = 30,
  mergeCutHeight    = 0.25,
  numericLabels     = FALSE,
  pamRespectsDendro = FALSE,
  saveTOMs          = FALSE,
  verbose           = 3
)

# How many modules were identified, and how many genes per module?
table(net$colors)
message(sprintf("Total modules identified: %d (excluding grey/unassigned)",
                sum(names(table(net$colors)) != "grey")))
```


```{r plot-dendrogram, fig.width=14, fig.height=6}
# Visualize the gene dendrogram with module color assignments
png(here("Analysis_Outputs", "10-5_wgcna_dendrogram.png"),
    width = 14, height = 6, units = "in", res = 300)
plotDendroAndColors(
  net$dendrograms[[1]],
  net$colors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels  = FALSE,
  hang          = 0.03,
  addGuide      = TRUE,
  guideHang     = 0.05,
  main          = "Gene dendrogram and module colors"
)
dev.off()

plotDendroAndColors(
  net$dendrograms[[1]],
  net$colors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels  = FALSE,
  hang          = 0.03,
  addGuide      = TRUE,
  guideHang     = 0.05,
  main          = "Gene dendrogram and module colors"
)
```

## 10.6 Module-Trait Correlation
```{r module-eigengenes}
# Extract module eigengenes — the first PC of each module's expression,
# summarizing the overall activity level of that module per sample.
MEs <- moduleEigengenes(expr_mat_t, colors = net$colors)$eigengenes
MEs <- orderMEs(MEs)

# Extract donor name from the pseudobulk sample ID (format: treatment_donor)
# and match to townlet estimates
sample_donors <- sub("^[^_]+_", "", rownames(MEs))

# Build the trait matrix — one row per pseudosample
# We use mean_beta as the primary sensitivity trait
trait_data <- data.frame(
  sample_id = rownames(expr_mat_t),
  donor     = extracted_donors
) %>%
  left_join(
    townlet %>% select(donor, mean_beta, mean_tau, sig_vs_baseline),
    by = "donor"
  ) %>%
  mutate(sig_numeric = as.numeric(sig_vs_baseline)) %>%
  select(sample_id, mean_beta, mean_tau, sig_numeric) %>%
  column_to_rownames("sample_id")

# Verify it looks right
head(trait_data)
sum(is.na(trait_data$mean_beta))  # should be 0

# Compute Pearson correlations between module eigengenes and traits
moduleTraitCor  <- cor(MEs, trait_data, use = "pairwise.complete.obs")
moduleTraitPval <- corPvalueStudent(moduleTraitCor, nrow(MEs))
```


```{r plot-module-trait-heatmap, fig.width=10, fig.height=12}
# Heatmap of module-trait correlations.
# Each cell shows r (top) and p-value (bottom).
# Modules with strong positive correlation to mean_beta are activated
# in lead-sensitive donors; negative correlation = activated in resistant donors.
textMatrix <- paste0(
  signif(moduleTraitCor, 2), "\n(",
  signif(moduleTraitPval, 1), ")"
)
dim(textMatrix) <- dim(moduleTraitCor)

# Plot
par(mar = c(6, 8.5, 3, 3))
labeledHeatmap(
  Matrix        = moduleTraitCor,
  xLabels       = c("Lead sensitivity\n(mean β)", "Tau effect\n(mean τ)", "Sig. vs baseline"),
  yLabels       = names(MEs),
  ySymbols      = names(MEs),
  colorLabels   = FALSE,
  colors        = blueWhiteRed(50),
  textMatrix    = textMatrix,
  setStdMargins = FALSE,
  cex.text      = 0.5,
  zlim          = c(-1, 1),
  main          = "Module-trait relationships"
)
dev.off()

par(mar = c(6, 8.5, 3, 3))
labeledHeatmap(
  Matrix      = moduleTraitCor,
  xLabels     = c("Lead sensitivity\n(mean β)", "Tau effect\n(mean τ)", "Sig. vs baseline"),
  yLabels     = names(MEs),
  ySymbols    = names(MEs),
  colorLabels = FALSE,
  colors      = blueWhiteRed(50),
  textMatrix  = textMatrix,
  setStdMargins = FALSE,
  cex.text    = 0.5,
  zlim        = c(-1, 1),
  main        = "Module-trait relationships"
)
```

## 10.7 Identifying Hub Genes in Significant Modules
```{r hub-genes}
# Identify modules significantly correlated with lead sensitivity (mean_beta)
# Use FDR-corrected p-values across all module-trait pairs
sig_modules <- data.frame(
  module  = rownames(moduleTraitCor),
  r       = moduleTraitCor[, "mean_beta"],
  p       = moduleTraitPval[, "mean_beta"]
) %>%
  mutate(
    padj      = p.adjust(p, method = "BH"),
    direction = ifelse(r > 0, "Sensitive", "Resistant")
  ) %>%
  filter(padj < 0.05) %>%
  arrange(padj)

print(sig_modules)

# For each significant module, extract the top hub genes —
# genes with the highest module membership (kME), which is the
# correlation of a gene's expression with the module eigengene.
# Hub genes are the most central and likely most functionally important
# members of the module.
kME <- signedKME(expr_mat_t, MEs, outputColumnName = "kME")

hub_gene_list <- list()

for (mod in sig_modules$module) {
  mod_color <- gsub("^ME", "", mod)
  mod_genes <- names(net$colors)[net$colors == mod_color]
  kme_col   <- paste0("kME", mod_color)
  
  hub_genes <- data.frame(
    gene_id = mod_genes,
    kME     = kME[mod_genes, kme_col]
  ) %>%
    arrange(desc(abs(kME))) %>%
    slice_head(n = 20) %>%
    left_join(ensg_to_symbol %>% select(gene_id, gene_name), by = "gene_id")
  
  hub_gene_list[[mod_color]] <- hub_genes
  message(sprintf("\nTop hub genes — %s module (r = %.2f, padj = %.3f):",
                  mod_color, sig_modules$r[sig_modules$module == mod],
                  sig_modules$padj[sig_modules$module == mod]))
  print(hub_genes %>% select(gene_name, kME))
}

# Save hub gene tables
hub_gene_df <- bind_rows(hub_gene_list, .id = "module")
write_csv(hub_gene_df, here("Analysis_Outputs", "10-7_wgcna_hub_genes.csv"))
```

```{r wgcna-enrichment, fig.width=10, fig.height=6}
# Run GSEA on hub genes for each significant module using kME as the ranking
# statistic — genes with highest kME are most central to the module
module_gsea_results <- list()

for (mod_color in names(hub_gene_list)) {
  
  r_beta        <- sig_modules$r[sig_modules$module == paste0("ME", mod_color)]
  kme_col       <- paste0("kME", mod_color)
  all_mod_genes <- names(net$colors)[net$colors == mod_color]
  
  # Annotate hub genes with direction relative to lead sensitivity
  # sign(kME) * sign(r_beta): +1 = high in sensitive, -1 = high in resistant
  hub_gene_list[[mod_color]] <- hub_gene_list[[mod_color]] %>%
    mutate(
      direction_vs_sensitivity = case_when(
        sign(kME) * sign(r_beta) > 0 ~ "High in sensitive donors",
        sign(kME) * sign(r_beta) < 0 ~ "High in resistant donors",
        TRUE ~ "neutral"
      )
    )
  
  # Build ranked list corrected for sensitivity direction:
  # positive rank = associated with lead sensitivity
  # negative rank = associated with lead resistance
  ranked <- kME[all_mod_genes, kme_col] * sign(r_beta)
  names(ranked) <- all_mod_genes
  ranked <- sort(ranked, decreasing = TRUE)
  ranked <- ranked[!is.na(ranked)]
  
  # Run GSEA against Reactome
  set.seed(42)
  gsea_res <- fgseaMultilevel(
    pathways = reactome.list,
    stats    = ranked,
    minSize  = 10,
    maxSize  = 500
  )
  
  module_gsea_results[[mod_color]] <- gsea_res %>%
    filter(padj < 0.05) %>%
    arrange(padj)
  
  message(sprintf("\n--- %s module (r_beta = %.2f, %s) --- %d significant pathways",
                  mod_color, r_beta,
                  sig_modules$direction[sig_modules$module == paste0("ME", mod_color)],
                  nrow(module_gsea_results[[mod_color]])))
  
  print(module_gsea_results[[mod_color]] %>%
          select(pathway, NES, padj) %>%
          head(15))
}

# Print hub gene directionality summary for top modules
for (mod_color in c("greenyellow", "brown", "red")) {
  r_beta <- sig_modules$r[sig_modules$module == paste0("ME", mod_color)]
  message(sprintf("\n--- ME%s hub genes (r_beta = %.2f) ---", mod_color, r_beta))
  print(
    hub_gene_list[[mod_color]] %>%
      select(gene_name, kME, direction_vs_sensitivity) %>%
      head(20)
  )
}

# Save results
saveRDS(module_gsea_results,
        file.path(output_dir, "10-7_wgcna_module_gsea.rds"))
write_csv(
  bind_rows(module_gsea_results, .id = "module"),
  here("Analysis_Outputs", "10-7_wgcna_module_pathway_enrichment.csv")
)

# Lollipop plots for top three modules
# Positive NES = enriched in sensitive donors
# Negative NES = enriched in resistant donors
for (mod_color in c("greenyellow", "brown", "red")) {
  
  if (nrow(module_gsea_results[[mod_color]]) == 0) {
    message(sprintf("No significant pathways for %s module", mod_color))
    next
  }
  
  direction_label <- sig_modules$direction[
    sig_modules$module == paste0("ME", mod_color)]
  r_beta <- sig_modules$r[sig_modules$module == paste0("ME", mod_color)]
  padj   <- sig_modules$padj[sig_modules$module == paste0("ME", mod_color)]
  
  top_up   <- module_gsea_results[[mod_color]] %>%
    filter(NES > 0) %>% slice_min(padj, n = 10)
  top_down <- module_gsea_results[[mod_color]] %>%
    filter(NES < 0) %>% slice_min(padj, n = 10)
  top_paths <- bind_rows(top_up, top_down)
  
  if (nrow(top_paths) == 0) next
  
  top_paths <- top_paths %>%
    mutate(
      pathway = gsub("REACTOME_", "", pathway),
      pathway = gsub("_", " ", pathway),
      pathway = str_to_title(pathway)
    )
  
  p <- ggplot(top_paths, aes(x = reorder(pathway, NES), y = NES)) +
    geom_segment(aes(xend = reorder(pathway, NES), yend = 0),
                 color = "gray60") +
    geom_point(aes(fill = NES > 0), pch = 21, size = 4, stroke = 0.8) +
    scale_fill_manual(
      values = c("TRUE"  = "#cc2132", "FALSE" = "#276bc4"),
      labels = c("TRUE"  = "High in sensitive donors",
                 "FALSE" = "High in resistant donors"),
      name   = NULL
    ) +
    coord_flip() +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    theme_classic() +
    theme(panel.border = element_rect(color = "black", fill = NA)) +
    labs(
      x        = NULL,
      y        = "Normalized Enrichment Score (NES)",
      title    = sprintf("ME%s module — Reactome pathway enrichment", mod_color),
      subtitle = sprintf("Lead %s module  |  r = %.2f, padj = %.2e",
                         direction_label, r_beta, padj)
    )
  
  print(p)
  
  ggsave(
    here("Analysis_Outputs",
         sprintf("10-7_wgcna_pathways_%s_module.png", mod_color)),
    plot   = p,
    width  = 11,
    height = 7,
    dpi    = 300,
    bg     = "white"
  )
}
```

```{r}
# Merge eigengenes with trait data and donor info for plotting
ME_wide <- MEs %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  mutate(donor = extracted_donors) %>%
  left_join(townlet %>% select(donor, mean_beta, sig_vs_baseline),
            by = "donor")

# Verify mean_beta is here before pivoting
head(ME_wide %>% select(sample_id, donor, mean_beta))

# Step 2: pivot only the ME columns
ME_plot_data <- ME_wide %>%
  pivot_longer(
    cols      = matches("^ME[a-z]"),
    names_to  = "module",
    values_to = "eigengene"
  ) %>%
  filter(module %in% sig_modules$module) %>%
  left_join(sig_modules %>% select(module, r, padj, direction),
            by = "module") %>%
  # join mean_beta back in from ME_wide after pivoting
  left_join(ME_wide %>% select(sample_id, mean_beta), by = "sample_id") %>%
  mutate(
    module_label = sprintf("%s\nr = %.2f, padj = %.2e", module, r, padj),
    module_label = factor(module_label,
                          levels = unique(module_label[order(abs(r),
                                                             decreasing = TRUE)]))
  )

head(ME_plot_data %>% select(sample_id, donor, mean_beta, eigengene, module))
ggplot(ME_plot_data, aes(x = eigengene, y = mean_beta)) +
  geom_point(aes(color = sig_vs_baseline), alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black",
              linewidth = 0.8, alpha = 0.15) +
  scale_color_manual(
    values = c("TRUE" = "#cc2132", "FALSE" = "grey60"),
    labels = c("TRUE" = "Sig. vs baseline", "FALSE" = "Not significant"),
    name   = NULL
  ) +
  facet_wrap(~ module_label, scales = "free_x", ncol = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40",
             linewidth = 0.4) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(size = 8, face = "bold"),
    panel.border     = element_rect(color = "black", fill = NA)
  ) +
  labs(
    x     = "Module eigengene expression",
    y     = "Lead sensitivity (mean β)",
    title = "Module eigengene vs lead sensitivity per pseudosample",
    subtitle = "Each point is one donor-treatment pseudosample (n=3 per donor) | Red = donor significantly responsive vs baseline"

  )

ggsave(here("Analysis_Outputs", "10-7_wgcna_eigengene_vs_sensitivity.png"),
       width = 16, height = 10, dpi = 300, bg = "white")

```





```{r save-wgcna-objects}
# Save all WGCNA outputs for reproducibility
saveRDS(net,             file.path(output_dir, "10-wgcna_network.rds"))
saveRDS(MEs,             file.path(output_dir, "10-wgcna_eigengenes.rds"))
saveRDS(moduleTraitCor,  file.path(output_dir, "10-wgcna_module_trait_cor.rds"))
write_csv(sig_modules,   here("Analysis_Outputs", "10-6_wgcna_significant_modules.csv"))

message("WGCNA analysis complete. Key outputs saved to Analysis_Outputs/.")
```




# Session Information

Recording your session information is a simple but important step for
reproducibility. It captures the exact R and package versions used,
allowing others (and your future self) to reproduce this analysis.

```{r session-info}
sessionInfo()
```
